A Switching Planner for Combined Task and Observation Planning

### 2010-11-29 code/subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/dtpddl.tex
Trimmed to eight awesome pages!

The modelling language of choice for planning in probabilistic
problems is the Probabilistic Planning Domain Definition
Language~\cite{younes:littman:04,younes:etal:2005}. PPDDL was used in
all 3 of the International Planning Competitions since 2004. A
variation on PDDL for describing domains with stochastic actions and
uncertain starting configurations, PPDDL is a declarative first-order
language that facilitates factored descriptions of domains and problem
instances. There are straightforward compilations from problems
expressed compactly in PPDDL to propositional representations amenable
to state-of-the-art planning procedures.

Because PPDDL cannot model domains that feature partial observability,
we develop an extension of that language, called Decision-Theoretic
(DT)PDDL. This can express probabilistic models of the sensing
consequences of acting, to quantitatively capture unreliability in
perception. That expressive power is achieved by incorporating
perceptual analogues of fluent, predicate, and action definitions. In
detail, we have declarations of state characterising predicate and
fluent symbols according to the PPDDL syntax. In addition, we allow
two other declarations, for perceptual predicates and fluents
respectively. For example, suppose our robot is tasked with exploring
{\em locations} in order to identify the whereabouts of a {\em
visual-object}. We must describe state and perceptual facts that model
the {\em true}, resp. perceived, locations of objects. In DTPDDL,
these declarations appear as:

@@

\noindent In a slight departure from PPDDL, we suppose that
$\pp{look-for-object}$ has an effect on the state. That is, its
execution incurs an instantaneous penalty, $3$, that corresponds to
the {\em cost} of performing the visual search.\footnote{In PPDDL {\em
reward} is a {\em reserved word}, and occurs in {\em increased} and
{\em decreased} terms in operator schemata. In that setting, {\em
reward} is interpreted as accumulated, whereas in DTPDDL it is
instantaneous.}

Completing the example, we model the sensing outcome that results from
instantiating the $\pp{look-for-object}$ operator:

@@

\noindent The interpretation of such an expression can be given
according to a {\em visitation} of terms. An atom is {\em visited} iff
its conjunctive parent is visited, and a conjunctive term is visited
iff all its immediate subterms are visited. A probabilistic term is
visited iff its conjunctive parent is visited, and exactly one of its
subterms, $T_i$, is visited. Each visitation of the root term
according to this recursive definition encapsulates a starting state,
along with the probability that it occurs. The former corresponds to the
union of all visited atoms, and the latter corresponds to the product
of $\prob_i$ entries on the visited subterms of probabilistic
elements. Making these ideas concrete, our example yields the
following flat state distribution:

@@

From hereon, in order to simplify the discussion (and implementation)
we shall restrict our attention to POMDPs with deterministic
actions. It is worthwhile noting that any POMDP with stochastic
actions can be compiled into an equivalent deterministic-action POMDP,
where all the original action uncertainty is expressed in the
starting-state distribution~\cite{ng:Jordan:2000}. In our setting,
that of finite-horizon planning, such a compilation yields a finite
flat representation of the original POMDP, only with deterministic
actions.

>>>>

The modelling language of choice for planning in probabilistic
problems is the Probabilistic Planning Domain Definition
Language~\cite{younes:littman:04,younes:etal:2005}. PPDDL has been
used in all three of the International Planning Competitions since
2004. A variation on PDDL for describing domains with stochastic
actions and uncertain starting configurations, PPDDL is a declarative
first-order language that facilitates factored descriptions of domains
and problems. There are straightforward compilations from problems
expressed in PPDDL to propositional representations amenable
to state-of-the-art planners.

Because PPDDL cannot model domains that feature partial observability,
we develop an extension we call {\em Decision-Theoretic
(DT)PDDL}. This can express probabilistic models of the sensing
consequences of acting, to quantitatively capture unreliability in
perception. That expressive power is achieved by incorporating
perceptual analogues of fluent, predicate, and action definitions. In
detail, we have declarations of state characterising predicate and
fluent symbols according to the PPDDL syntax. In addition, we allow
two other declarations, for perceptual predicates and fluents
respectively. For example, suppose our robot is tasked with exploring
{\em locations} in order to identify the whereabouts of a {\em
visual-object}. We must describe state and perceptual facts that model
the {\em true}, resp. perceived, locations of objects. In DTPDDL,
these declarations appear as:

@@

\noindent In a slight departure from PPDDL, we suppose that
$\pp{look-for-object}$ has an effect on the state. That is, its
execution incurs an instantaneous penalty, $3$, that corresponds to
the {\em cost} of performing the visual search.\footnote{In PPDDL {\em
reward} is a {\em reserved word}, and occurs in {\em increased} and
{\em decreased} terms in operator schemata. In that setting, reward is
accumulated, whereas in DTPDDL it is instantaneous.}

Completing the example, we model the sensing that results from
instantiating the $\pp{look-for-object}$ operator:

@@

\noindent The interpretation of such an expression can be given
by a {\em visitation} of terms: An atom is {\em visited} iff
its conjunctive parent is visited, and a conjunctive term is visited
iff all its immediate subterms are visited. A probabilistic term is
visited iff its conjunctive parent is visited, and exactly one of its
subterms, $T_i$, is visited. Each visitation of the root term
according to this recursive definition encapsulates a starting state,
along with the probability that it occurs. The former corresponds to the
union of all visited atoms, and the latter corresponds to the product
of $\prob_i$ entries on the visited subterms of probabilistic
elements. Making these ideas concrete, our example yields the
following flat state distribution:

@@

From hereon, to simplify the discussion (and implementation)
we shall restrict our attention to POMDPs with deterministic
actions. We note that POMDPs with stochastic
actions can be compiled into equivalent deterministic-action POMDPs,
where all the original action uncertainty is expressed in the
starting-state distribution~\cite{ng:Jordan:2000}. In our setting,
that of finite-horizon planning, such a compilation yields a finite
flat representation of the original POMDP, only with deterministic
actions.

### 2010-11-29 code/subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/intro.tex
Trimmed to eight awesome pages!

Planning for agents acting in partially observable environments with
stochastic actions is extremely computationally
expensive~\cite{mdp-complexity}. Even relatively small partially
observable Markov decision problems (POMDPs), represented by tens of
propositional variables, produce state spaces far larger than existing
optimal or close to optimal planners can handle. For the easier
problem of planning in completely observable stochastic domains, an
alternative approach has been to use classical sequential planners and
replan when actions have unplanned-for outcomes~\cite{yoon:etal:2007}.
Execution monitoring tracks properties of the state that determine
whether a new plan is necessary. For partially observable domains,
only probabilistic information about such properties will, in general,
be available.

@@

@@

Our switching system is domain independent, taking domain and problem
descriptions in a first-order declarative language we have developed,
called the decision-theoretic planning domain definition language
(DTPDDL). From those descriptions of POMDPs, we automatically
construct a deterministic model for the sequential planner. That model
includes operators which correspond to assumptions about the values of
imperfectly known state variables. Assumptions scheduled by the
sequential system are used to propose a pragmatic abstract
belief-state space to the decision-theoretic system, and to modify the
reward function, so that system might pursue sensing related to those
assumptions.

@@

Our switching planner gives us some important advantages in our
robotic domain. First, the significant source of uncertainty in the
domain is the unreliability of observations, and this approach is
particularly suited to problems that combine a deterministic task
planning problem with observations as the decision-theoretic planner
is only invoked to determine a characteristic of the underlying
state. Secondly, the replanning makes the system very robust to
changing objectives and the discovery of new facts about the world,
both of which are features of our domain.

>>>>

Planning for agents acting in partially observable environments with
stochastic actions is extremely computationally
expensive~\cite{mdp-complexity}. Even relatively small partially
observable Markov decision problems (POMDPs), represented by tens of
propositional variables, have state spaces far larger than existing
optimal or close to optimal planners can handle. For the easier
problem of planning in completely observable stochastic domains, an
alternative has been to use classical sequential planners and
replan when actions have unplanned-for outcomes~\cite{yoon:etal:2007}.
Execution monitoring tracks properties of the state that determine
whether further planning is required. For partially observable
domains, only probabilistic information about such properties will, in
general, be available.

@@

%% Switching gives us some important advantages in our robotic
%% domain. First, the significant source of uncertainty in the domain is
%% the unreliability of observations, and this approach is particularly
%% suited to problems that combine a deterministic task planning problem
%% with observations as the decision-theoretic planner is often invoked
%% to determine a characteristic of the underlying state. Secondly,
%% replanning makes our system robust to changing objectives and the
%% discovery of new facts about the world.

@@

Our system is domain independent, taking domain and problem
descriptions in a first-order declarative language we have developed,
called the decision-theoretic planning domain definition language
(DTPDDL). From these descriptions of POMDPs, we automatically
construct a deterministic model for the sequential planner. That model
includes actions which correspond to making assumptions about the values of
imperfectly known state variables. Assumptions scheduled by the
sequential system are used to propose a pragmatic abstract
belief-state space to the decision-theoretic system, and to modify the
reward function, so that system might pursue sensing related to those
assumptions.

@@

The switching planner provides important advantages in our
mobile robot domain. First, the significant source of uncertainty in the
domain is the unreliability of observations and this approach is
particularly suited to problems that combine a deterministic task
planning problem with observations because the decision-theoretic planner
is only invoked to deal with state uncertainty.
%determine a characteristic of the underlying state. 
Secondly, replanning makes the system robust to
changing objectives and discoveries about the world,
both of which feature in our domain.

### 2010-11-29 code/subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/prop-dtp.tex
typo in value function
Trimmed to eight awesome pages!

We describe the partially observable propositional probabilistic
planning problem, with costs and rewards. We model a process
state $\state$ as the set of propositions that are true of the
state. Notationally, we have $\state \subseteq
\props$. The underlying process
dynamics are modelled in terms of a finite set of {\em probabilistic
STRIPS operators}~\cite{boutilier:1996} $\stochActions$ over
state-characterising propositions $\propositions$.
%%
Notationally, we say an action $\stochAction \in \stochActions$ is
applicable if its precondition $\poss(\stochAction)$, a set of
propositions, are satisfied in the current state -- i.e., $\poss(\stochAction) \subseteq
\state$. We denote $\mu_{\stochAction}(\detAction_i)$ the probability that
nature chooses a deterministic STRIPS effect $\detAct_i$, and for
all \stochAction\ we require
$\sum_{\detAction_i \in \detActions(\stochAction)}
\mu_{\stochAction}(\detAction_i) = 1$.

@@

Finally, we make a number of fairly standard assumptions. First, that
action execution and sensing occurs instantaneously, and that only one
action can be applied at a plan-step. Second, it can arise that in
some propositional states an action is applicable, and in others it is
not. Moreover, a belief-state can assign non-zero probability to
states where that applicability holds, and states where it does
not. We differ
from~\citeauthor{younes:littman:04}~(\citeyear{younes:littman:04}),
because we forbid the execution of actions that are not applicable in
any state of the current belief.  Also,
unlike~\citeauthor{hoffmann:brafman:2006}~(\citeyear{hoffmann:brafman:2006}),
we do incorporate a PPDDL-like default semantics, treating an action
\stochAction\ executed at $\state_i$ as if it has no effect at
states $\state$ if $\poss(\stochAction) \not\subseteq
\state$.

@@

Whereas until now we have considered a POMDP model that is factored in
terms of propositions and percepts, in order to discuss utilities and
policies it is convenient to consider the underlying decision process
in a flat format. This underlying decision process is given by the
tuple $\langle \states, \bstate_0, \actions, \transProb, \reward,
\observations, \obsDist \rangle$. Here $\bstate_0$ is the initial
belief-state, \states\ is the finite set of reachable propositional
states, \actions\ is the finite set of actions, and \observations\ is
the finite set of reachable observations (i.e., perceptual states).
Where $\state, \state' \in \states$, $a \in \actions$, from $\mu$ we
have a state transition function $\transProb(\state, \action,
\state')$ giving the probability of a transition from state $s$ to
$s'$ if $a$ is applied. For any $\state$ and $\action$ we have
$\sum_{\state' \in \states} \transProb(\state, \action, \state') = 1$.
%%
Function $\reward:\states \times \actions \to \Re$ is a bounded real
valued reward function. Therefore a finite positive constant $c$
exists so that for all $\state \in \states$ and $\action \in
\actions$, $|\reward(\state, \action)| < c$. We model costs as
negative rewards.
%%
From $\psi$ we have that for each $\state \in \states$ and action
$\action \in \actions$, an observation $\observ \in \observations$ is
generated independently according to a probability distribution
$\obsDist(\state, \action)$. We denote $\obsDist_\observ(\state,
\action)$ the probability of getting observation $\observ$ in state
$\state$. For $\state$ and $\action$ we have $\sum_{\observ \in
\observations} \obsDist_\observ(\state, \action) = 1$.

In the POMDP model successive state estimation is available by
application of Bayes' rule.  Taking the current belief $\bstate$ as
the {\em prior}, and supposing action $\action$ is executed with
perceptive outcome $\observ$, then the probability that we are in
$\state$ in the successive belief-state $\bstate'$ is given by:

>>>>

We describe the partially observable propositional probabilistic
planning problem, with costs and rewards. We model a process
state $\state$ as the set of propositions that are true of the
state. Notationally, we have $\state \subseteq
\props$. The underlying process
dynamics are modelled in terms of a finite set of {\em probabilistic
STRIPS operators}~\cite{boutilier:abstraction} $\stochActions$ over
state-characterising propositions $\propositions$.
%%
We say an action $\stochAction \in \stochActions$ is
applicable if its precondition $\poss(\stochAction)$, a set of
propositions, are satisfied in the current state -- i.e., $\poss(\stochAction) \subseteq
\state$. We denote $\mu_{\stochAction}(\detAction_i)$ the probability that
nature chooses a deterministic STRIPS effect $\detAct_i$, and for
all \stochAction\ we require
$\sum_{\detAction_i \in \detActions(\stochAction)}
\mu_{\stochAction}(\detAction_i) = 1$.

@@

Finally, we make some fairly standard assumptions. First, that
action execution and sensing occur instantaneously, and that only one
action can be applied per plan-step. Second, 
that
%it can arise that in
%some propositional states an action is applicable, and in others it is
%not. Moreover, a belief-state can assign non-zero probability to
%states where that applicability holds, and states where it does
%not. We differ
%%in situations where an action exists that is applicable in some state 
%% with non-zero probability, and not applicable in another, we differ
%from~\citeauthor{younes:littman:04}~(\citeyear{younes:littman:04}),
%because 
%%in that
%we forbid the execution of the action.
%actions that are not applicable in
%any state of the current belief.  
%Also,
unlike~\citeauthor{hoffmann:brafman:2006}~(\citeyear{hoffmann:brafman:2006}),
we incorporate a PPDDL-like default semantics, treating an action
\stochAction\ executed at $\state_i$ as if it has no effect at
states $\state$ if $\poss(\stochAction) \not\subseteq
\state$.

% Mucked around with this, then realised the two sentences were
% contradictory. If we forbid actions applicable in one state but not
% another in the belief state, then clearly we aren't using a
% PPDDL-like semantics.
%% Doubled comments are stuff I added, then removed.

@@

Until now we have discussed the POMDP in terms of propositions and
percepts. In order to address belief revision and utility it is
convenient to consider the underlying decision process in a flat
format. This is given by the tuple
$\langle \states, \bstate_0, \actions, \transProb, \reward,
\observations, \obsDist \rangle$. Here, $\bstate_0$ is the initial
belief-state, \states\ is the finite set of reachable propositional
states, \actions\ is the finite set of actions, and \observations\ is
the finite set of reachable observations (i.e., perceptual states).
Where $\state, \state' \in \states$, $a \in \actions$, from $\mu$ we
have a state transition function $\transProb(\state, \action,
\state')$ giving the probability of a transition from state $s$ to
$s'$ if $a$ is applied. For any $\state$ and $\action$ we have
$\sum_{\state' \in \states} \transProb(\state, \action, \state') = 1$.
%%
Function $\reward:\states \times \actions \to \Re$ is a bounded real
valued reward function. Therefore a finite positive constant $c$
exists so that for all $\state \in \states$ and $\action \in
\actions$, $|\reward(\state, \action)| < c$. We model costs as
negative rewards.
%%
From $\psi$ we have that for each $\state \in \states$ and action
$\action \in \actions$, an observation $\observ \in \observations$ is
generated independently according to a probability distribution
$\obsDist(\state, \action)$. We denote $\obsDist_\observ(\state,
\action)$ the probability of getting observation $\observ$ in state
$\state$. For $\state$ and $\action$ we have $\sum_{\observ \in
\observations} \obsDist_\observ(\state, \action) = 1$.

Successive state estimation is by application of Bayes'
rule.  Taking the current belief $\bstate$ as the {\em prior}, and
supposing action $\action$ is executed with perceptive outcome
$\observ$, the probability that we are in $\state$ in the
successive belief-state $\bstate'$ is:

### 2010-11-29 code/subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/relatex.tex
Trimmed to eight awesome pages!

There have been a number of developments recently moving towards
planning under uncertainty using systems that were intended for
sequential planning in deterministic problems.  Notably for example,
\system{FFR$_a$}~\cite{yoon:etal:2007}, the winning entry from the
probabilistic track of the 2004 International Planning Competition.
In the continual paradigm, \system{FFR$_a$} uses the fast satisficing
procedure \system{FF}~\cite{hoffmann:nebel:2001} to compute sequential
plans and corresponding execution traces.
%%
More computationally expensive approaches in this vein combine
sampling strategies on valuations over {\em runtime variables} with
deterministic planning procedures. The outcome is typically a more
robust sequential plan~\cite{yoon:etal:2008}, or contingent
plan~\cite{majercik:2006}.

@@

Also leveraging deterministic planners in problems that feature
uncertainty, \system{Conformant-FF}~\cite{hoffmann:brafman:2006} and
$T_0$~\cite{palacios:geffner:2009} demonstrate how conformant planning
---i.e., sequential planning in unobservable worlds--- can be modelled
as a deterministic problem, and therefore solved using sequential
systems. In this conformant setting, advances have been towards
compact representations of beliefs amenable to existing best-first
search planning procedures, and lazy evaluations of beliefs. Most
recently this research thread has been extended to contingent planning
in fully observable non-deterministic
environments~\cite{albore:etal:2009}.
%%
The continual planning system that motivated our
project~\cite{brenner:nebel:jaamas09} also has this characteristic,
and has been applied in completely observable domains, particularly
those featuring multiple communicating agents. The use of knowledge
operators in domains allows plans that act to gain knowledge, but the
approach assumes that such actions are deterministic and reliable, an
assumption that we relax in this work.

>>>>

There have been a number of recent papers on
planning under uncertainty using systems that were intended for
sequential planning in deterministic problems,  for example,
\system{FFR$_a$}~\cite{yoon:etal:2007}, the winning entry from the
probabilistic track of the 2004 International Planning Competition.
In the continual paradigm, \system{FFR$_a$} uses the fast satisficing
procedure \system{FF}~\cite{hoffmann:nebel:2001} to compute sequential
plans and corresponding execution traces.
%%
More computationally expensive approaches in this vein combine
sampling strategies on valuations over {\em runtime variables} with
deterministic planning procedures. The outcome is typically a more
robust sequential plan~\cite{yoon:etal:2008}, or contingent
plan~\cite{majercik:2006}.

@@

Also leveraging deterministic planners in problems that feature
uncertainty, \system{Conformant-FF}~\cite{hoffmann:brafman:2006} and
$T_0$~\cite{palacios:geffner:2009} demonstrate how conformant planning
---i.e., sequential planning in unobservable worlds--- can be modelled
as a deterministic problem, and therefore solved using sequential
systems. In this conformant setting, advances have been towards
compact representations of beliefs amenable to existing best-first
search planning procedures, and lazy evaluations of beliefs. Most
recently this research thread has been extended to contingent planning
in fully observable non-deterministic
environments~\cite{albore:etal:2009}.
%%
The continual planning system that motivated our
project~\cite{brenner:nebel:jaamas09} also has this characteristic,
and has been applied in completely observable domains, particularly
those featuring multiple communicating agents. The use of knowledge
operators in domains allows plans that act to gain knowledge, but the
approach assumes that such actions are deterministic and reliable, an
assumption that we relax.

### 2010-11-29 code/subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/switching.tex
Trimmed to eight awesome pages!

For sequential planning the operators from the POMDP model are
available with semantics that accommodate the above abstraction as
follows. In the deterministic model a proposition $\prop$ can be
thought to have a ternary interpretation at a state \state, as either
{\em true}, written $\prop\in\state$, {\em false}, written
$\prop\not\in\state$, or {\em unspecified}, with a slight abuse of
notation written $\prop\#\in\state$. For example, in $\pp{DORA}$
$(=(\pp{is-in}~\pp{box})\pp{office})\#\in\state_0$. For POMDP actions
$\action\in\actions$ in the deterministic model, if
$\prop \in \poss(\action)$, or if $\prop$ is the subject of a positive
or negative effect of $\action$, then $\action$ is not applicable
in \state\ if $\prop\#\in\state$. In other words, an assumptive action
must be applied to make a truth assignment to \prop, and then that
truth assignment could be modified by the DTPDDL domain operators.
%%
Lastly, with regards to switching, we have that if $\action$ has a
precondition that is not true with probability $1$ in the underlying
belief $\bstate$, then immediately before it is executed blindly, the
system switches to a contingent session, discussed in the following
section.

>>>>

For sequential planning the operators from the POMDP model are
available with semantics that accommodate the above abstraction as
follows: In the deterministic model a proposition $\prop$ can be
thought to have a ternary interpretation at a state \state, as either
{\em true}, written $\prop\in\state$, {\em false}, written
$\prop\not\in\state$, or {\em unspecified}, with a slight abuse of
notation written $\prop\#\in\state$. For example, in $\pp{DORA}$
$(=(\pp{is-in}~\pp{box})\pp{office})\#\in\state_0$. For POMDP actions
$\action\in\actions$ in the deterministic model, if
$\prop \in \poss(\action)$, or if $\prop$ is the subject of a positive
or negative effect of $\action$, then $\action$ is not applicable
in \state\ if $\prop\#\in\state$. In other words, an assumptive action
must be applied to make a truth assignment to \prop, and then that
truth assignment could be modified by the DTPDDL domain operators.
%%
Lastly, we have that if $\action$ has a
precondition that is not true with probability $1$ in the underlying
belief $\bstate$, then immediately before it is executed blindly, the
system switches to a contingent session, discussed in the following
section.
